# Tools & Advice for Instructors

```{r advice_setup, include=FALSE}

# Load libraries
library(fivethirtyeight)
library(psych)
library(skimr)

# A rediculous hack to get the unicode characters in skim_df's to show up on windows
# Basically we pretend we're in China, and everything is OK.
LC_CTYPE <- Sys.getlocale("LC_CTYPE")
if (tolower(.Platform$OS.type) == "windows") {
  Sys.setlocale("LC_CTYPE", "Chinese")
}

# Tailor skimr
skim_with(numeric=list(complete=NULL, p0=NULL, p25=NULL, p50=NULL, p75=NULL, p100=NULL), 
          integer=list(complete=NULL, p0=NULL, p25=NULL, p50=NULL, p75=NULL, p100=NULL), 
          logical=list(complete=NULL), 
          character=list(complete=NULL), 
          factor=list(complete=NULL, ordered=NULL))

```

## Best Practices

### Commands vs. Buttons

Though we strongly encourage you to use R through R Studio, we also encourage you and your students to rely on commands instead of button presses where possible. For instance, it is possible to save plots by exporting them from the Plots pane in R Studio, but it is also possible to save them with commands such as `pdf()` or `png()`. There are several benefits to prioritizing commands over button presses. First, it has the practical effect of reinforcing a command-line approach to programming your analyses, which may be new for you and is very likely new for your students. Second, and perhaps more importantly, it encourages reproducibility. If your students document all of the actions that they took in completing an assignment by having a command for each one, it will be easier for you to make sure that they did everything on their own, correct their habits where needed, and find the source of an error. It will also help you to reproduce your own work from semester to semester if your assignments are coded in R.

### Naming Objects

There are relatively few constraints to naming objects in R compared to other statistical software packages. We make several recommendations here for good naming practices.

The "naming conventions" of a particular programming language refer to the formatting guidelines for object names. There are no agreed-upon naming conventions in R, but a good overview can be found [here](https://journal.r-project.org/archive/2012-2/RJournal_2012-2_Baaaath.pdf). In general, the best advice is to pick a convention that appeals to you and use it consistently. We recommend the "underscore_separated" scheme. Because period separation can sometimes refer to a different process in R, we do not recommend using the "period.separated" scheme.

Regardless of the convention, it is also important to choose object names that are informative. That is, someone with little knowledge of the script, including some future version of yourself, should be able to get a general idea of what data an object contains based on the name. For instance, simply naming all versions of a data frame `data`, `data_2`, `data_3`, etc., is not informative because it is unclear what changes have been made between versions. 

### Scripting

More than just a collection of commands, the script is an excellent tool for both both reproducibility and documentation of an analysis. It is therefore worthwhile to teach students good habits for writing scripts.

First and foremost, scripts should include *all of the commands that you need to conduct an analysis, and none of the commands that you don't*. That is, you should be able to run all of the lines in a script in order and get the same result each time (with the obvious exception of things like drawing random samples). To that end, all the commands necessary for an analysis should be in the proper sequence, and deny commands that are not necessary for the analysis but kept for posterity should be "commented out" by adding the `#` symbol at the start of the line.

Second, scripts should be well-documented. Recall that you can add comments to any line by preceding the comment with the `#` symbol. Comments are an efficient way to explain to someone unfamiliar with the script what the intention is for each section of code. Ample documentation is not only a good programming habit to teach your students, it will also likely be very useful in helping you decipher their code.

### Formatting Requirements for Assignments

We encourage you to give your students clear formatting guidelines for assignments that require a significant amount of R programming to complete. The specific requirements that you decide on should be based on (1) the extent to which you want them to leave the class with generalizeable programming skills, and (2) the amount of detail that you would like from your students in terms of their code. 

Asking students to print the contents of their console will allow you to see all of the commands that they ran, the order in which they ran them, as well as the output and any error messages, but it will be quite a lot of information for you to sort through yourself. At the other extreme, asking for only the formatted results in a Word document makes it difficult to judge the quality of their code, which may be a primary goal of the assignment. A nice compromise is to ask for an electronic version of their script with all the necessary commands so that you can reproduce their results. This can either act as the main document or as a supplement to a formatted report.

It has also become increasingly popular for instructors to provide students with prepared R Markdown (.Rmd) files with fill-in-the-blank style prompts for students to enter their code. This produces a nicely formatted document with their code and output, which may be easier for you administratively. However, this can also obscure the coding process for students and limit their independence. For this reason, we recommend against using prepared R Markdown files, and encourage you to teach students how to create and maintain their own scripts instead.

## Error Messages

All messages printed to the console in red text are meant to get your attention, but they differ in urgency. A message with `Error` printed at the beginning means that there was a problem that prevented the command from being evaluated, and you will not be able to run that command until you fix the problem. A message with `Warning` printed at the beginning means that the command *was* evaluated, but it might not have worked the way you intended, or R might have made some decision for you in order to execute the command. All other messages can generally be taken as updates on the status of your command.

If you receive an error message that is unclear, the best way to determine the cause is to Google it. A list of the most common errors is provided here for convenience.

### There is no output from a command, and there is a `+` in the console instead of a `>`. {-}

The user has not completed the command, most likely by forgetting to close a quotation (or being inconsistent with the use of single vs. double quotations) or a set of parentheses. You can either add the end quotation or parenthesis if it is the final character, or press the escape button and start again.

### `Error: unexpected ')' in '___'` {-}

The user has included one too many end parentheses.

### `Error: Object '___' not found` {-}

The user has tried to call an object that is not in the environment. This may be because there is a typo in the name of the object, or because the object was never assigned to a label with the `<-` operator.

### `Error in ___ : non-numeric argument to binary operator` {-}

The user is most likely trying to conduct a mathematical operation on data that is of a non-numeric class, such as character or factor. If the data should be numeric, try overwriting it with `as.numeric()`.

### `Error in ___(x) : could not find function '___'` {-}

The user is attempting to use a function that is not currently available in R, either because there is a typo in the name of the function or because the function's package has not been loaded in this session.

### `Error in ___ : cannot open the connection` {-}

The user is attempting to read or write a file at a location on the hard drive that does not exist, likely due to a typo in the file path or file name.

### `Error in ___ : undefined columns selected** or **Error in ___: subscript out of bounds` {-}

The user is attempting to index an area of an object that does not exist, for instance, the ninth column of a seven-column data frame.

### `Error: unexpected symbol in '___'` {-}

R cannot interpret part of the command, likely because the user has forgotten a comma between function arguments.

## Data Sets

This section includes an overview of some selected data sets that may be useful for in-class demonstrations or assignments. The first section includes data sets that are "built-in" as part of the base R `datasets` package, whereas the remaining sections review data sets from other selected packages that need to be installed. Of course, this list is non-exhaustive. The packages used here have more data sets than those we have chosen to highlight, and there are many more interesting data sets in other packages we have not highlighted. So, feel free to explore more options.

To see all of the data sets available from all *loaded* packages, including those are built-in with base R, enter the following command:

```{r data list, eval=FALSE}
data()
```

You should see a list of all the available data sets with a brief summary of each, separated by the package they belong to. To see a list of datsets in *all* installed packages (loaded or not), run the command `data(package = .packages(all.available = TRUE))`.

### How to Handle `NA` Values

Most of the data sets below include at least some missing data, marked as `NA`, in at least one variable. Because, as you know, missing data is a problem that is sure to appear in the wild for students who continue on a research path, you may choose to teach students how to handle `NA` values themselves. Alternatively, you may choose to remove `NA` values on your own for simplicity, either by removing them from the data set beforehand (in which case the data would need to be saved as a .CSV file for the students to read in later), or providing students with code to run before beginning the assignment. Regardless, this section reviews how to handle `NA` values.

As mentioned previously, `NA` values can be identified with the `is.na()` function. For instance, imagine that we want to identify the missing quantitative SAT scores in the `sat.act` data set. We can first call `is.na()` on the variable to create a logical vector marking each `NA` as `TRUE`:

```{r is.na identify}
is.na(sat.act$SATQ)
```

This logical vector can then be used to index the variable in question:

```{r is.na index}
sat.act[is.na(sat.act$SATQ), 'SATQ']
```

The previous command returns the `NA` values because we have asked R to show us all of the values in `sat.act$SATQ` variable that are coded as `NA`. However, we are often more interested in the values that are NOT missing than those that are. To find the *non*-missing values, we can reverse the logical vector by adding the negation operator, `!`, to the beginning of the call to `is.na()`:

```{r is.na index negate}
sat.act[!is.na(sat.act$SATQ), 'SATQ']
```

Now we've asked R to show us all of the values in the `sat.act$SATQ` variable that are NOT coded as `NA`.

To index the entire `sat.act` data frame so that *all* columns are returned, not just the quantitative SAT scores, use bracket notation without specifying any columns:

```{r is.na index negate full call, eval=FALSE}
sat.act[!is.na(sat.act$SATQ), ]
```
```{r is.na index negate full output, echo=FALSE}
head(sat.act[!is.na(sat.act$SATQ), ])
```

Now we have the full `sat.act` data set, but only those rows where the quantitative SAT score is not `NA`. To be complete, we recommend saving this subset as a new object:

```{r is.na subset}
sat.act_noNA <- sat.act[!is.na(sat.act$SATQ), ]
```

Note that you can also overwrite `NA` values, in cases where it is methodologically appropriate. For instance, perhaps you know that the students with missing quantitative SAT scores are treated as having a score of 0. You could replace the `NA` values with 0 as follows:

```{r replace na}

# First save a new version of the data
sat.act_NA0 <- sat.act

# Recode NA values as 0 in new version
sat.act_NA0[is.na(sat.act$SATQ), 'SATQ'] <- 0

```

Now all missing quantitative SAT scores will be treated as a 0 in future mathematical operations, using this new version of the data. Importantly, the original data frame is also preserved.

### Built-In

There are several small data sets included in base R that are useful for quick demos. Since they are part of base R, it is not necessary to load these packages; just type and enter their name in the console to view them.

Note that the descriptions and summary statistics shown for each dataset were computed using the `skim()` function from the `skimr` pacakge.

#### iris {-}

The `iris` data set gives the lengths and widths of the sepals and petals of 50 irises from each of three different species:

```{r iris summary, echo=FALSE}
skim(iris)
```

#### warpbreaks {-}

The `warpbreaks` data set gives the number of breaks in each of 54 looms of yarn, varied across two types of wool and three levels of tension:

```{r warpbreaks summary, echo=FALSE}
skim(warpbreaks)
```

#### mtcars {-}

The `mtcars` data set gives a range of numeric and categorical measurements of 32 cars, including MPG, number of cylinders, horsepower, weight, and transmission type:

```{r mtcars summary, echo=FALSE}
skim(mtcars)
```

#### ToothGrowth {-}

The `ToothGrowth` data set gives the tooth lengths of 60 guinea pigs, varied across three vitamin C dosage levels and two dosage delivery methods:

```{r ToothGrowth summar, echo=FALSE}
skim(ToothGrowth)
```

### FiveThirtyEight

The `fivethirtyeight` package includes data sets used for articles written on Nate Silver's website for data geeks, [FiveThirtyEight](https://fivethirtyeight.com/). From an instructional standpoint, these data sets are useful not only for being particularly engaging but because there is a published article associated with each one.

To make the data sets available, install and load the `fivethirtyeight` package:

```{r load fivethirtyeight, eval=FALSE}
install.packages('fivethirtyeight')
library(fivethirtyeight)
```

#### bechdel {-}

The `bechdel` data set is a great all-purpose data set with both within- and between-subject continuous and categorical variables. Based on the [Bechdel Test](https://en.wikipedia.org/wiki/Bechdel_test) for female representation in film, this data set includes 1,794 movies from 1970-2013 along with their test outcome, budget, and gross profits:

```{r bechdel summary, echo=FALSE}
skim(bechdel)
```

#### fandango {-}

The `fandango` data set shows ratings from 146 movies across the Fandango, IMDB, Rotten Tomatoes, and Metacritic websites. It is particularly useful for demonstrating correlation:

```{r fandango summary, echo=FALSE}
skim(fandango)
```

#### comma_survey {-}

The `comma_survey` data set has Likert ratings as well as demographic information collected from 1,129 respondents to a survey on the Oxford comma:

```{r oxford_comma summary, echo=FALSE}
skim(comma_survey)
```

#### avengers {-}

The `avengers` data set includes the death and revival status for 173 Marvel comic book characters. It largely includes count and categorical data:

```{r avengers summary, echo=FALSE}
skim(avengers)
```

#### bob_ross {-}

The `bob_ross` data set is a great tool for demonstrating count data and tables. For each of 403 episodes, the data set marks the presence or absence of a series of possible elements in Bob's painting. Useful combinations include tables of trees, lakes, and snowy mountains:

```{r bob_ross summar, echo=FALSE}
skim(bob_ross)
```

### psych

In addition to a range of new functions, the `psych` package includes data sets intended to demonstrate common statistical analyses for psychological research.

To make the data sets available, install and load the `psych` package:

```{r ref.label="load_psych", eval=FALSE}
```

#### Tal.Or {-}

The `Tal.Or` data set is from Study 2 of Tol-Or, Cohen, Tasfati, & Gunther (2010), which used an experimental manipulation to test the effects of media on the attitudes of 123 respondents, measured with Likert scales:

```{r Tal.Or summary, echo=FALSE}
skim(Tal.Or)
```

#### affect {-}

The `affect` data set consists of responses from 330 participants to state and trait mood scales given before and after viewing (1) a Frontline documentary about the liberation of a concentration camp, (2) the horror movie Halloween, (3) a National Geographic nature film about the Serengeti, or (4) the comedy movie Parenthood:

```{r affect summary, echo=FALSE}
skim(affect)
```

#### epi {-}

The `epi` data set includes responses to the Eysenck Personality Inventory from 3570 subjects:

```{r epi summary, echo=FALSE}
skim(epi)
```

#### sat.act {-}

The `sat.act` data set includes self-reported SAT and ACT scores along with demographic information from 700 respondents, making it useful for within- or between-subject comparisons:

```{r sat.act summary, echo=FALSE}
skim(sat.act)
```

```{r, include=FALSE}
Sys.setlocale("LC_CTYPE", LC_CTYPE) # Restore LC_CTYPE to original value
```


### Simulating your own Data Sets

If none of the available data sets satisfy your needs, or if you simply want to get creative, it is possible to simulate your own data set by drawing random samples from a distribution. This can be a great method of designing engaging data sets. It also ensures that you will have a nice "clean" data set, with no unwanted `NA` values, balanced or unbalanced designs as needed, and null and significant effects as desired.

To demonstrate how to simulate a numeric variable, the following code simulates income for 1,000 families from a population in which income is normally distributed around $50,000 with a standard deviation of $5,000:

```{r numeric sample demo}
sim_income <- rnorm(n=1000, mean=50000, sd=5000)
hist(sim_income)
```

To demonstrate how to simulate a categorical variable, the following code simulates grades for 50 students in a pass/fail course in which there is a 50% chance of passing in the population:

```{r categorical sample demo}
sim_grades <- rbinom(n=50, size=1, prob=.5)
sim_grades <- factor(sim_grades, levels=0:1, labels=c('FAIL', 'PASS'))
barplot(table(sim_grades))
```

Other potentially useful "population" distributions include `rbeta()`, `rbinom()`, `rchisq()`, `rexp()`, `rf()`, `rgamma()`, `rlogis()`, `rlnorm()`, `rpois()`, `rt()`, and `runif()`. A full list can be found [here](https://en.wikibooks.org/wiki/R_Programming/Probability_Distributions). See the help documentation for each to see which parameter values can be specified.

#### Reproducibility {-}
The R session information when compiling this book is shown below:
```{r reproducibility}
sessionInfo()
```
